Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-10-29 manuscript.Rmd
more writing

---
author:
  - name: Daniel W. Hill
    affiliation: University of Georgia
    address: >
      First line
      Second line
    email: dwhill@uga.edu
    url: http://spia.uga.edu/directory/faculty_staff/hill-daniel
  - name: Nils W. Metternich
    affiliation: University College London
  - name: Andreas Beger
    affiliation: Duke University
  - name: Michael D. Ward
    affiliation: Duke University
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model. The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{R}"]
  plain:     [keywords, not capitalized, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

```{r data}
library("spduration")
#data(bscoup)
library("foreign")
bscoup <- read.dta("data/BelkinSchoferTable4.dta")
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year")
```

\normalsize
The \code{add\_duration} function takes as input a data frame with a binary response variable (in this example, "coup"") that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named "duration") for each spell as well as a binary variable ("atrisk") that is constant across the spell and indicates whether the spell ended in failure. These are then used as response variables for the formula in the `spdur` function. The \code{spdur} function is the primary function in the package and produces a regression model object of class \code{spdur} which can then be used with further methods.

Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data. We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation.

```{r models, include = FALSE}
weib_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup, silent = TRUE)

@@

Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values. One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects. 
```{r fit-stats}
AIC(weib_model)
AIC(loglog_model)
BIC(weib_model)
BIC(loglog_model)
```

The objects produced by \code{spdur} can also be used with functions in the \code{xtable} package to produce \LaTeX\ code for tables of results. The \code{xtable} function used on an \code{spdur} object will produce code such as that used to create Table \ref{loglog_table}.

```{r summary-table, include=FALSE}
library("xtable")
tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
              label="loglog_table")
print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
```

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} 
\label{loglog_table}
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 3.30 & 0.14 & 0.00 \\ 
  milreg & -1.53 & 0.15 & 0.00 \\ 
  instab & -0.22 & 0.03 & 0.00 \\ 
  regconf & 3.04 & 2.22 & 0.17 \\ 
  log(alpha) & -0.52 & 0.07 & 0.00 \\ 
  (Risk Intercept) & -14.61 & 9.02 & 0.11 \\ 
  couprisk & 9.11 & 4.37 & 0.04 \\ 
  wealth & 3.51 & 1.75 & 0.05 \\ 
  milreg.1 & 21.76 & 0.24 & 0.00 \\ 
  rwar & -6.37 & 5.38 & 0.24 \\ 
  regconf.1 & -20.22 & 54.04 & 0.71 \\ 
  samerica & -3.31 & 3.28 & 0.31 \\ 
  camerica & -6.46 & 2.57 & 0.01 \\ 
   \hline
\end{tabular}
\end{table}

@@

Finally, we demonstrate how to evaluate a model's out of sample predictions. We begin by splitting the data into training and test sets, using the last year of data (1999) as the test set and omitting incomplete observations from the test set. 
```{r}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year")
```

```{r}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year")
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

\normalsize
The \code{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.

```{r}
weib_model2   <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, silent = TRUE)

@@

Foo

@@

\normalsize
The \code{train},\code{test}, and \code{pred} arguments are used to specify data frames for the training set, test set, and forecasting set. We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set. Here we use \code{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population. The test set predictions are stored in the model object as \code{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \code{separationplot} package.

```{r oos-sepplots}
library("separationplot")

obs_y <- coup_test$coup[complete.cases(coup_test)]

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
\end{center}
\end{figure}

```{r}
glm_model2    <- glm(
  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
    samerica + camerica,
  data = coup_train, family = "binomial"
)

@@

@@

Many outcomes in the area that we work in, political science and political violence more specifically, are rare events. Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable. Figure \ref{rates} shows a few examples for published research that models binary outcomes. @fearon2003ethnicity is a very widely cited study of civil war onset that uses yearly observations of all major countries, and @beger2015ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.

In the language of machine learning, we are dealing with highly imbalanced classes, although our data are also highly structured with clear dependence over space and time in countries and societies that may experience political violence. This is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling]. Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics.

Split-population duration models are not only appealing in this technical sense, but they also match the logic or intuituion many analysts of politics use when they distinguish long-term risk factors from more fleeting triggering causes. The example we have used, @belkin2003toward, is a particularly clear illustration of how well the language of theorists maps onto the model intuition.

>>>>

---
author:
  - name: Daniel W. Hill
    affiliation: University of Georgia
    address: >
      First line
      Second line
    email: dwhill@uga.edu
    url: http://spia.uga.edu/directory/faculty_staff/hill-daniel
  - name: Nils W. Metternich
    affiliation: University College London
  - name: Andreas Beger
    affiliation: Duke University
  - name: Michael D. Ward
    affiliation: Duke University
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model. The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{R}"]
  plain:     [keywords, not capitalized, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

```{r data}
library("spduration")
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year", ongoing = FALSE)
```

The \code{add\_duration} function takes as input a data frame with a binary response variable--"coup"--that measures the occurrence of the event, or failure, recorded over discrete time periods. We assume that the data frame is in a cross-sectional time-series format that consists of multiple observations over time of a number of subjects, in our case countries.

Within the framework of duration modeling, these data conceptually consist of "spells", which are repated observations of a unit from the time they enter the data (left-censoring) or after they experienced a failure event, until the next failure event or end of observation (right censoring). Table \ref{tab-ex} shows how a single country could have two spells that consist of the observations for that country from 1960, when our data start, to a failure event in say 1980, when a new spell starts and continues until the next failure event or data end. Country "B" on the other hand experienced no coups and all observation for that country from 1960 to 2000 make up one spell.

\begin{table}
\begin{center}
\caption{Example data frame with select duration variables}
\label{tab-ex}
\begin{tabular}{lrrrrr} \toprule
Country & Year & Coup & Spell ID & Duration \\ \midrule
A & 1960 & 0 & 1 & 0  \\ 
A & \ldots & 0 & 1 & \ldots \\
A & 1980 & 1 & 1 & 20 \\ \midrule
A & 1981 & 0 & 2 & 0 \\
A & \ldots & 0 & 2 & \ldots \\
A & 2000 & 0 & 2 & 19  \\ \midrule
B & 1960 & 0 & 3 & 0 \\
B & \ldots & \ldots & 3 & \ldots \\ 
B & 2000 & 0 & 3 & 40 \\ \bottomrule
\end{tabular}
\end{center}
\end{table}

The \code{add_duration} functions identifies spells in the data and returns the data frame with several additional variables needed for estimation. The most important among these are a counter of elapsed time in a spell, "duration". It also codes that a spell was at risk ("atrisk==1") if it ended in a failure event, and not at risk otherwise.   These are then used as response variables for the formula in the \code{spdur} function.

Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data and will try to convert the \code{tID} input to class "Date" given the provided argument in \code{freq}. Another important question is how to handle consecutive 1's in the response variable provided. This is controlled with the \code{ongoing} argument. In some cases like civil war occurence, the *y* variable records all years during which a country experienced a civil war. Here \code{ongoing} should be set to true so that the models try to predict the onset of the civil war, but disregard ongoing conflicts (\code{failure} is set to \code{NA} for these cases, dropping them from analysis). With ongoing set to false, successive 1's in *y* are treated as distinct and new failures, and kept in the analysis. This makes sense for discrete, short-term events like coups. Countries can experience distinct coups in successive years.

The \code{spdur} function is the primary function in the package and produces a regression model object of class \code{spdur} which can then be used with further methods. We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation.

```{r models, include = TRUE, cache = TRUE}
weib_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup, silent = TRUE)

@@

Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values. These can be exported by calling \code{xtable} directly on the "spdur" object to produce Table \ref{loglog-table}.

```{r summary-table, include = TRUE, eval = FALSE}
library("xtable")
tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
              label="loglog_table")
print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
```

One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.

```{r fit-stats}
matrix(c(
  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
```

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} 
\label{loglog_table}
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
  milreg & -1.13 & 0.21 & 0.00 \\ 
  instab & -0.09 & 0.02 & 0.00 \\ 
  regconf & -2.52 & 2.16 & 0.24 \\ 
  log(alpha) & -0.45 & 0.06 & 0.00 \\ 
  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
  couprisk & 0.59 & 0.32 & 0.07 \\ 
  wealth & -0.36 & 0.28 & 0.20 \\ 
  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
  rwar & -0.52 & 0.94 & 0.58 \\ 
  regconf.1 & -5.43 & 5.62 & 0.33 \\ 
  samerica & 2.09 & 1.45 & 0.15 \\ 
  camerica & -0.39 & 0.73 & 0.59 \\ 
   \hline
\end{tabular}
\end{table}

@@

## Out-of-sample testing

Finally, we demonstrate how to evaluate a model's out of sample predictions. We will use data from 1996 on as the test set, and the prior data for training. The \code{add_duration} function retrospectively codes the risk variable based on how a spell ended, and we therefore need to take care in how we add the duration variables for each data set. For the training data we need to subset the training set first so that coups in the tests set don't influence the risk coding in the training data.

```{r, eval = FALSE}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year", ongoing = FALSE)
```

```{r, include = FALSE}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year", ongoing = FALSE)
```

For the test set it is better to add the duration variables and *then* subset the test set. Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information. To do this we need to build the duration variables first:

```{r, include = FALSE}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year", ongoing = FALSE)
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

```{r, eval = FALSE}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year", ongoing = FALSE)
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

Now we can fit new models using the training data, and calculate predictions from these models for the test data, using \code{predict(..., newdata = coup_test)}.

```{r estimate train models, cache = TRUE}
weib_model2   <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, silent = TRUE)

@@

@@

Since we are predicting for data that is not contained in the "spdur" model objects, we have to use the \code{separationplot} functions directly from the package to produce Figure \ref{oos-sepplots}.

```{r oos-sepplots, eval = FALSE}
library("separationplot")
obs_y <- coup_test[complete.cases(coup_test), "coup"]

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
\caption{Out-of-sample separation plots}
\label{oos-sepplots}
\end{center}
\end{figure}

Lastly, here is a comparision to regular logistic regression.

```{r logistic regression}
glm_model2    <- glm(
  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
    samerica + camerica,
  data = coup_train, family = "binomial"
)

@@

```{r, eval = FALSE, include = FALSE}
glm2_test_p   <- predict(glm_model2, newdata = coup_test, type = "response")
glm2_test_p   <- glm2_test_p[!is.na(glm2_test_p)]

obs_y <- coup_test[complete.cases(coup_test), "coup"]

library("separationplot")
separationplot(glm2_test_p,   obs_y)

brier(obs_y, glm2_test_p)
auc_roc(obs_y, glm2_test_p)
auc_roc(obs_y, weib2_test_p)
auc_roc(obs_y, loglog2_test_p)
auc_pr(obs_y, glm2_test_p)
auc_pr(obs_y, weib2_test_p)
auc_pr(obs_y, loglog2_test_p)
```

@@

Many outcomes in the area that we work in, political science and political violence more specifically, are rare events. Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable. Figure \ref{rates} shows a few examples for published research that models binary outcomes. @fearon2003ethnicity is a very widely cited study of civil war onset that uses yearly observations of all major countries, and @beger2014ensemble is an example of the positive rates when moving to monthly data for a similar set of countries. The positive rates range from 5 to less than 0.2\% of all data points.

In the language of machine learning, we are dealing with highly imbalanced classes, this is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling]. Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics that are an important part of the non-independent data political scientists commonly use.

Split-population duration models are not only appealing in a technical sense, but they also match the logic or intuituion many analysts of politics use when they distinguish long-term risk factors from more fleeting triggering causes. The example we have used, @belkin2003toward, is a particularly clear illustration of how well the language of theorists maps onto the model intuition.
